---
title: "analysis_code"
author: "Valérian Fraisse"
date: "2026-03-17"
output: html_document
---

Composing with acoustic niches: A laboratory soundscape evaluation of sound installation reactivity

Valérian Fraisse

The present code is the analysis code used on the listening tests reported 
in the journal article 'Composing with acoustic niches: A laboratory soundscape evaluation 
of sound installation reactivity' by Fraisse V., Schütz N., Vincent C., Wanderley M., Guastavino C., and Misdariis N., submitted to npj Acoustics. 


The code is not functional per se as participants' data from the listening tests
can't be shared online due to the nature of consent obtained
from participants as per the ethics approval obtained from the McGill Research
Ethics Board and the Sorbonne University Research Ethics Commitee. 

Code sections are ordered according to the paper's sections they are 
refering to.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
# Libraries

# Libraries for data manipulation
library(dplyr) # Pipes and data manipulation
library(tidyr) # Data tidying
library(reshape2) # melt, utility
library(stringr) # string manipulation, utility
library(plyr) # ddply, utility
library(purrr) # map, utility
library(forcats) # tiddy up categorical variables, utility

# Libraries for statistical analysis
library(psych) # PCA and other stats
library(car) # Levene's test
library(rstatix) # Wilcoxon effect size
library(nlme) # Linear mixed models
library(MANOVA.RM) # Repeated measures MANOVA
library(mvnormtest) # Shapiro-Wilk multivariate normality tests
library(lmerTest) # Linear mixed models with Satterthwaite's method for degrees of freedom

# function to compute post-hoc, pairwise Wilcoxon tests
pairwise_wilcox = function(df, var_cols, cond_col, p_method, ci=FALSE, nboot=1000)
{
  result <- data.frame()
  for (col in var_cols) {
    pairwise.wilcox.test(df[[col]],
                         df[[cond_col]],
                         paired = TRUE, 
                         p.adjust.method = p_method,
                         exact=TRUE) -> temp_w
    
    wilcox_effsize(df,
                   as.formula(paste(colnames(df[col]), 
                                    colnames(df[cond_col]), 
                                    sep="~")), 
                   paired = TRUE,                
                   ci = ci,
                   nboot = nboot) -> temp_es
    
    temp_vals <- na.omit(melt(temp_w$p.value, 
                              value.name=paste("pval", p_method, sep="_")
    ))
    
    temp_vals$item <- colnames(df[col])
    temp_vals$effsize <- round(temp_es$effsize, 2)
    
    if (ci) {
      temp_vals$effsize_cl <- temp_es$conf.low
      temp_vals$effsize_ch <- temp_es$conf.high
    }
    
    temp_vals$magnitude <- temp_es$magnitude
    
    result <- na.omit(bind_rows(result, temp_vals))
  }
  return(result)
}

# Function to test normality of pairwise differences
test_difference_normality <- function(data, scales, condition_col, scenario_name) {
  print(paste("Testing difference normality for:", scenario_name))
  
  conditions <- unique(data[[condition_col]])
  condition_pairs <- combn(conditions, 2, simplify = FALSE)
  
  for(pair in condition_pairs) {
    print(paste("Comparing:", pair[1], "vs", pair[2]))
    
    cond1_data <- data[data[[condition_col]] == pair[1], scales]
    cond2_data <- data[data[[condition_col]] == pair[2], scales]
    
    # Calculate differences for each scale
    for(scale in scales) {
      if(nrow(cond1_data) > 0 && nrow(cond2_data) > 0) {
        participants1 <- data[data[[condition_col]] == pair[1], "participant"]
        participants2 <- data[data[[condition_col]] == pair[2], "participant"]
        common_participants <- intersect(participants1, participants2)
        
        if(length(common_participants) > 3) {  # Need sufficient sample size
          diff_scores <- cond1_data[participants1 %in% common_participants, scale] - 
                        cond2_data[participants2 %in% common_participants, scale]
          
          # Shapiro-Wilk test on differences
          if(length(diff_scores) > 3 && length(diff_scores) <= 5000) {
            shapiro_result <- shapiro.test(diff_scores)
            print(paste("  ", scale, "- Shapiro p-value:", round(shapiro_result$p.value, 4)))
          }
        }
      }
    }
  }
}

# Function to test ANOVA residual normality
test_anova_residual_normality <- function(data, scales, condition_col, subject_col, scenario_name) {
  print(paste("Testing ANOVA residual normality for:", scenario_name))
  
  for(scale in scales) {
    print(paste("Scale:", scale))
    
    # Fit linear mixed model to get residuals (equivalent to RM ANOVA)
    formula_str <- paste(scale, "~", condition_col)
    
    tryCatch({
      model <- lme(as.formula(formula_str), 
                   random = as.formula(paste("~ 1 |", subject_col)), 
                   data = data,
                   na.action = na.omit)
      
      residuals <- residuals(model)
      
      if(length(residuals) > 3 && length(residuals) <= 5000) {
        shapiro_result <- shapiro.test(residuals)
        print(paste("  Shapiro p-value:", round(shapiro_result$p.value, 4)))
        
        ks_result <- ks.test(residuals, "pnorm", mean(residuals), sd(residuals))
        print(paste("  KS p-value:", round(ks_result$p.value, 4)))
      }
      
    }, error = function(e) {
      print(paste("  Error fitting model for", scale, ":", e$message))
    })
  }
}
```

## Tests for preliminary assumptions
```{r assumptions, include=TRUE,  fig.width=10, fig.height=10}
# Univariate normality
for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  print(sce)
  for (col in 11:18) {
    temp <- as.numeric(unlist(filter(all.data, scenario == sce)[col]))
    if (shapiro.test(temp)$p < .05) {
      print(colnames(all.data[col]))
      print(shapiro.test(temp))    
    }
  }}

# Multivariate normality
for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  print(sce) 
  all.data %>%
    filter(scenario == sce) %>% 
    # select(so_trafic:so_autre) %>% 
    select(sc_agreable:sc_emergence) %>%
    t() %>% 
    mshapiro.test() %>% 
    print()
}

# Homogeneity of variance
for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  print(sce)
  for (col in 11:18) {
    print(colnames(all.data[col]))
    temp <- as.numeric(unlist(all.data[col]))
    print(leveneTest(filter(all.data, scenario == sce)[[col]] 
                     ~ filter(all.data, scenario == sce)[[19]], 
                     data = filter(all.data, scenario == sce)))
  }
}

# Homogeneity of covariance matrices
# For soundscape scales
for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  print(paste("Scenario:", sce))
  scenario_data <- all.data %>% filter(scenario == sce)
  
  # Box's M test for soundscape scales
  boxM_result <- boxM(scenario_data[, c("sc_agreable", "sc_apaisante", "sc_niveau", 
                                       "sc_caractere", "sc_appropriee", "sc_coherente",
                                       "sc_variee", "sc_emergence")], 
                     scenario_data$condition)
  print(boxM_result)
}

# For sound source scales  
for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  print(paste("Scenario:", sce))
  scenario_data <- all.data %>% filter(scenario == sce)
  
  # Box's M test for sound source scales
  boxM_result <- boxM(scenario_data[, c("so_trafic", "so_passants", "so_nature", 
                                       "so_music", "so_autre")], 
                     scenario_data$condition)
  print(boxM_result)
}

# Normality of group differences

# For soundscape scales
soundscape_scales <- c("sc_agreable", "sc_apaisante", "sc_niveau", 
                      "sc_caractere", "sc_appropriee", "sc_coherente",
                      "sc_variee", "sc_emergence")

for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  scenario_data <- all.data %>% filter(scenario == sce)
  test_difference_normality(scenario_data, soundscape_scales, "condition", sce)
}

# For sound source scales  
source_scales <- c("so_trafic", "so_passants", "so_nature", "so_music", "so_autre")

for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  scenario_data <- all.data %>% filter(scenario == sce)
  test_difference_normality(scenario_data, source_scales, "condition", sce)
}

# Normality of residuals

# Test for each scenario - soundscape scales

for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  scenario_data <- all.data %>% filter(scenario == sce)
  test_anova_residual_normality(scenario_data, soundscape_scales, "condition", "participant", sce)
}

# Test for sound source scales  
source_scales <- c("so_trafic", "so_passants", "so_nature", "so_music", "so_autre")

for (sce in c("Rush hour", "Afterwork", "Quiet time")) {
  scenario_data <- all.data %>% filter(scenario == sce)
  test_anova_residual_normality(scenario_data, source_scales, "condition", "participant", sce)
}

```

## Loudness validation
```{r loudness}
lmm_loudness <- lmerTest::lmer(sc_niveau ~ loudness + scenario + (1 | participant),
                      data = all.data)
summary(lmm_loudness) 
```

## 2.1 Comparing Baseline Scenarios

```{r analysis_scenarios, include=TRUE,  fig.width=12, fig.height=12}
# Sound source dominance
# 1 - RM MANOVA
scenarios.data %>% 
  multRM(formula = cbind(so_trafic, so_passants, so_nature,
                         so_music, so_autre) ~ scenario, 
         subject = "participant", 
         within = "scenario",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 6:10) {
  items <- append(items, colnames(scenarios.data[col]))
  RM(data=scenarios.data, 
     formula = scenarios.data[[col]] ~ scenario, 
     subject = "participant", 
     within = "scenario",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

scenarios_sources_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
sources_wilcox <- pairwise_wilcox(scenarios.data,
                                c(6,7), 
                                22, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)

# Soundscape scales
# 0 - Filtering Baseline scenarios
scenarios.data = filter(
  all.data, 
  condition == "Baseline scenario"
)

# 1 - RM MANOVA
scenarios.data %>% 
  multRM(formula = cbind(sc_agreable, sc_apaisante, sc_appropriee,
                         sc_variee, sc_coherente, sc_caractere,
                         sc_niveau, sc_emergence) ~ scenario, 
         subject = "participant", 
         within = "scenario",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 11:18) {
  items <- append(items, colnames(scenarios.data[col]))
  RM(data=scenarios.data, 
     formula = scenarios.data[[col]] ~ scenario, 
     subject = "participant", 
     within = "scenario",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

scenarios_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
scenarios_wilcox <- pairwise_wilcox(scenarios.data,
                                c(11,12,14,16,17), 
                                22, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)
```

## 2.2 Rush hour Scenario

```{r analysis_soundscape_vs_scenarios_rushhour, include=TRUE,  fig.width=12, fig.height=12}
# 0 - Filtering scenario
rush_hour.data = filter(
  all.data, 
  scenario == "Rush hour"
)

# Sound source dominance
# 1 - RM MANOVA
rush_hour.data %>% 
  multRM(formula = cbind(so_trafic, so_passants, so_nature, so_music, so_autre) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 6:10) {
  items <- append(items, colnames(rush_hour.data[col]))
  RM(data=rush_hour.data, 
     formula = rush_hour.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

rush_hour_sources_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
rush_hour_sources_wilcox <- pairwise_wilcox(rush_hour.data,
                                c(7,8,9,10), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)
# Soundscape scales
# 1 - RM MANOVA
rush_hour.data %>% 
  multRM(formula = cbind(sc_agreable, sc_apaisante, sc_appropriee,
                         sc_variee, sc_coherente, sc_caractere,
                         sc_niveau, sc_emergence) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 11:18) {
  items <- append(items, colnames(rush_hour.data[col]))
  RM(data=rush_hour.data, 
     formula = rush_hour.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

rush_hour_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS,
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
rush_hour_wilcox <- pairwise_wilcox(rush_hour.data,
                                c(13, 14,15,16,18), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)
```

## 2.3 Afterwork Scenario

```{r analysis_soundscape_vs_scenarios_afterwork, include=TRUE,  fig.width=13, fig.height=12}
# 0 - Filtering scenario
afterwork.data = filter(
  all.data, 
  scenario == "Afterwork"
)

# Sound sources
# 1 - RM MANOVA
afterwork.data %>% 
  multRM(formula = cbind(so_trafic, so_passants, so_nature, so_music, so_autre) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 6:10) {
  items <- append(items, colnames(afterwork.data[col]))
  RM(data=afterwork.data, 
     formula = afterwork.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

afterwork_sources_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
afterwork_sources_wilcox <- pairwise_wilcox(afterwork.data,
                                c(6,7,8,9,10), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)

# Soundscape scales
# 1 - RM MANOVA
afterwork.data %>% 
  multRM(formula = cbind(sc_agreable, sc_apaisante, sc_appropriee,
                         sc_variee, sc_coherente, sc_caractere,
                         sc_niveau, sc_emergence) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 11:18) {
  items <- append(items, colnames(afterwork.data[col]))
  RM(data=afterwork.data, 
     formula = afterwork.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

afterwork_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
afterwork_wilcox <- pairwise_wilcox(afterwork.data,
                                c(13,14,15,16), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)
```

## 2.4 Quiet time Scenario

```{r analysis_soundscape_vs_scenarios_quiettime, include=TRUE,  fig.width=12, fig.height=12}

# 0 - Filtering scenario
quiet_time.data = filter(
  all.data, 
  scenario == "Quiet time"
)

# Sound sources
# 1 - RM MANOVA
quiet_time.data %>% 
  multRM(formula = cbind(so_trafic, so_passants, so_nature, so_music, so_autre) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 6:10) {
  items <- append(items, colnames(quiet_time.data[col]))
  RM(data=quiet_time.data, 
     formula = quiet_time.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

quiet_time_sources_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
quiet_time_sources_wilcox <- pairwise_wilcox(quiet_time.data,
                                c(6,7,8,9,10), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)

# Soundscape scales
# 1 - RM MANOVA
quiet_time.data %>% 
  multRM(formula = cbind(sc_agreable, sc_apaisante, sc_appropriee,
                         sc_variee, sc_coherente, sc_caractere,
                         sc_niveau, sc_emergence) ~ condition, 
         subject = "participant", 
         within = "condition",
         iter = 10000, 
         resampling = "WildBS", seed = 1234) %>% 
  summary()

# 2 - Follow-up semi-parametric RM Anova
p_vals <- c()
items <- c()
ATS <- c()
for (col in 11:18) {
  items <- append(items, colnames(quiet_time.data[col]))
  RM(data=quiet_time.data, 
     formula = quiet_time.data[[col]] ~ condition, 
     subject = "participant", 
     within = "condition",
     iter = 10000, 
     resampling = "WildBS", seed = 1234) -> temp_RM 
  ATS <- append(ATS, temp_RM$ATS[1])
  p_vals <- append(p_vals, temp_RM$resampling[2])
}

quiet_time_RM_ANOVA <- data.frame("item" = items, 
                             "ATS" = ATS, 
                             "p-value_BH" = p.adjust(p_vals, "BH"))

# 3 - Post-hoc Wilcoxon tests
quiet_time_wilcox <- pairwise_wilcox(quiet_time.data,
                                c(11,12,14,16,18), 
                                19, 
                                "BH",
                                ci=TRUE,
                                nboot=5000)
```
